Regularised parameter selection in the non-stationary Hawkes process

### 2017-05-03 _paper_output/chapter_sparse_hawkes.tex
refs

    The log-likelihood function for a given realisation of the data is can
be found from the usual likelihood for Poisson data of a given intensity
\cite{ozaki_etimating_1979}

@@

Using the method of maximum likelihood we estimate our parameter vector
\(\theta:=(\mu, \phi)\) by choosing the estimates to maximise the
log-likelihood,
\[\hat{\theta}_\pi(N) = \o{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

@@

In practice, we choose
\[\psi_j(t) := \mathbb{I}_{[\tau_{j},\tau_{j+1})}(t)\] for some lattice
\(\{\tau_j\},\) which will correspond to the assumption that "sparse"
triggers will have localised support in time; we are looking for short
"spikes".

@@

\[\begin{aligned}
  { \scriptstyle \frac{\partial}{\partial\kappa_i} } L_\pi(N; \theta)&=-\int_0^T \sum_{t_i\lt t}\phi_i(t-t_i) dt + \sum_{0\leq t_i\leq T} \frac{\sum_{t_k\lt t_i}\phi_j(t_i-t_k)}{\lambda_\theta(t_i)} -\frac{\partial}{\partial\kappa_j}\pi(\theta)
\end{aligned}\]

@@

\[\begin{aligned}
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\pi\|\boldsymbol \omega_\theta(t)\|_1 &= { \scriptstyle \frac{\partial}{\partial\omega_j} }\pi \int_{\tau_{j-1}}^{\tau_j} \left|\omega_j\right|dt\\
&=\pi\Delta\tau_j\operatorname{sgn} \omega_j\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\lambda_\theta(t|\mathcal{F}_t) &=  \Delta\tau_j\mathbb{I}_{[\tau_{j-1},\tau_j)}(t)\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} } L_\pi(\mathbf t;\theta) &=-\Delta\tau_j-\pi\Delta\tau_j\operatorname{sgn} \omega_j + \sum_{\tau_{j-1}\leq t_i\leq \tau_j} \frac{1}{\lambda_\theta(t_i)}
\end{aligned}\]

@@

The traditional means of performing model selection in Hawkes-type
models e.g. \cite{ogata_estimation_1983}, uses \emph{degrees-of-freedom}
penalties \cite{efron_how_1886}. Specifically, Akaike's
\cite{akaike_likelihood_1981} Information Criterion, or \emph{AIC}, is
widely popular.

@@

\[\lim_{\pi\to\infty}\hat{\theta}_\pi'(\mathbf t) =\hat{\theta}(\mathbf t)\]

@@

\[L_\pi(\mathbf t;\mu,\phi):=-\int_0^T\lambda(t;\mu,\phi)dt + \int_0^T\log \lambda(t;\mu,\phi) dN_t - \pi\|\boldsymbol \omega_\theta(t)\|_1\]

@@

    We also, when notationally convenient, use the equivalent representation
as the left-continuous function that increments at each arrival time: \[
N(t)-N(s):= \left|\{s:s\in \mathbf{t}\,\mathrm{and}\,s\leq t\}\right| \,\mathrm{ for }\, t\geq s
\]

>>>>

    The log-likelihood function for a given realisation of the data is can
be found from the usual likelihood for Poisson data of a given intensity
\cite{ozaki_estimating_1979}

@@

Using the method of maximum likelihood we estimate our parameter vector
\(\theta:=(\mu, \phi)\) by choosing the estimates to maximise the
log-likelihood,
\[\hat{\theta}_\pi(N) = \oo{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

@@

In practice, we choose
\[\psi_j(t) := \bb{I}_{[\tau_{j},\tau_{j+1})}(t)\] for some lattice
\(\{\tau_j\},\) which will correspond to the assumption that "sparse"
triggers will have localised support in time; we are looking for short
"spikes".

@@

\[\begin{aligned}
  { \scriptstyle \frac{\partial}{\partial\kappa_i} } L_\pi(N; \theta)&=-\int_0^T \sum_{t_i\lt t}\phi_i(t-t_i) \dd t + \sum_{0\leq t_i\leq T} \frac{\sum_{t_k\lt t_i}\phi_j(t_i-t_k)}{\lambda_\theta(t_i)} -\frac{\partial}{\partial\kappa_j}\pi(\theta)
\end{aligned}\]

@@

\[\begin{aligned}
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\pi\|\v \omega_\theta(t)\|_1 &= { \scriptstyle \frac{\partial}{\partial\omega_j} }\pi \int_{\tau_{j-1}}^{\tau_j} \left|\omega_j\right|\dd t\\
&=\pi\Delta\tau_j\oo{sgn} \omega_j\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\lambda_\theta(t|\mathcal{F}_t) &=  \Delta\tau_j\bb{I}_{[\tau_{j-1},\tau_j)}(t)\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} } L_\pi(\v t;\theta) &=-\Delta\tau_j-\pi\Delta\tau_j\oo{sgn} \omega_j + \sum_{\tau_{j-1}\leq t_i\leq \tau_j} \frac{1}{\lambda_\theta(t_i)}
\end{aligned}\]

@@

The traditional means of performing model selection in Hawkes-type
models e.g. \cite{ogata_estimation_1983}, uses \emph{degrees-of-freedom}
penalties \cite{efron_how_1986}. Specifically, Akaike's
\cite{akaike_likelihood_1981} Information Criterion, or \emph{AIC}, is
widely popular.

@@

\[\lim_{\pi\to\infty}\hat{\theta}_\pi'(\v t) =\hat{\theta}(\v t)\]

@@

\[L_\pi(\v t;\mu,\phi):=-\int_0^T\lambda(t;\mu,\phi) \dd t + \int_0^T\log \lambda(t;\mu,\phi) \dd N_t - \pi\|\v \omega_\theta(t)\|_1\]

@@

    We also, when notationally convenient, use the equivalent representation
as the left-continuous function that increments at each arrival time: \[
N(t)-N(s):= \left|\{s:s\in \v{t}\,\text{ and }\,s\leq t\}\right| \,\mathrm{ for }\, t\geq s
\]
